\vspace{1.0cm} \Large \bf Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''

### 2016-04-27 mSystems/mSys_MAIN.tex
Some work in mSys MAIN (fix paper geometry) and Results (v0.0.8)
On branch master
	modified:   mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_results.tex
	modified:   results/hist/antibiotics_h_patE_pre_AbsFreqPlot.pdf

% release codenumber (update it here!)
\newcommand\releasecode{0.0.7}

@@

% text layout
%\usepackage[letterpaper]{geometry}
%\geometry{textwidth=16.25cm} % 15.25cm for single-space, 16.25cm for double-space
%\geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space

@@

\title{
	\vspace*{-16mm} % Helps to include the corresponding author footnote in the title page
	\singlespacing
	\begin{flushleft}
		\texttt{\large Title:} \\
	\end{flushleft}
	\vspace*{2mm}
	Health and disease imprinted in the time variability of the human microbiome\\
	\vspace*{6mm}
	\begin{flushleft}
		\texttt{\large Running title:} \\
	\end{flushleft}
	\vspace*{0mm}
	Microbiota, are you sick?
	\vspace*{4mm}
	}

>>>>

% release codenumber (update it here!)
\newcommand\releasecode{0.0.8}

@@

% text layout
\usepackage[letterpaper]{geometry}
\geometry{textwidth=16.25cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22.5cm} % 22cm for single-space, 22.5cm for double-space

@@

\title{
	\vspace*{0mm} % May help to include the corresponding author footnote in the title page
	\singlespacing
	\begin{flushleft}
		\texttt{\large Title:} \\
	\end{flushleft}
	\vspace*{2mm}
	Health and disease imprinted in the time variability of the human microbiome\\
	\vspace*{6mm}
	\begin{flushleft}
		\texttt{\large Running title:} \\
	\end{flushleft}
	\vspace*{0mm}
	Microbiota, are you sick?
	\vspace*{4mm}
	}

### 2016-04-27 mSystems/mSys_results.tex
Some work in mSys MAIN (fix paper geometry) and Results (v0.0.8)
On branch master
	modified:   mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_results.tex
	modified:   results/hist/antibiotics_h_patE_pre_AbsFreqPlot.pdf

>>>>

%%%
\subsection*{Specific results}

\subsubsection*{Fit Plots}
For each and every dataset included in the study, an unweighted fit (see Material and Methods section for details) and a X-weighted fit (detailed in Material and Methods subsection) have been calculated for standard deviations versus the mean values for each bacterial genus monitored in time. Figure \ref{fig:main1} showed the X-weighted fit for samples from a healthy subject (patient A, top) and from a subject diagnosed with irritable bowel syndrome (patient P2, bottom) studied in our lab\cite{IBS}, while Figure \ref{fig:fits} shows the corresponding unweighted fits. Additionally, for the unweighted fit, a complete residues analysis was performed, and a 4-in-1 figure was generated as shown in Figure \ref{fig:unwRes}, corresponding to patient A (top plot in Figure \ref{fig:main1} and Figure \ref{fig:fits}). Among other tests, it allows to check for normality and homoscedasticity of the residues and, therefore, to further evaluate the goodness of the fit.

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_LLR_LOG.pdf}
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_P1_metatranscriptomes_family_stdVSmean_LLR_LOG.pdf}
	\caption{Log plots of unweighted fits corresponding to the datasets shown in Figure \ref{fig:main1}}
	\label{fig:fits}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_LLR_RES.pdf}
	\caption{Residues analysis plot corresponding to the unweighted fit for patient A of the IBS study\cite{IBS}. The top-left subplot is a simple residues plot. The top-right subplot is a Normal quantiles plot with linear fitting (value of coefficient of determination is provided). The bottom-left subplot shows an accumulated ESS (Explained Sum of Squares) plot. Finally, the bottom-right subplot is a residues Normal histogram plot. This set of subplots allows to check for normality and homoscedasticity of the residues.}
	\label{fig:unwRes}
\end{figure}

\subsubsection*{Histogram Plots} 
\CC\ generates two related histogram plots, absolute frequencies plot and zero relative frequency plot. The former is useful to visually assess the validity of the time points in terms of the accumulated absolute frequency of the elements (taxa), since absolute frequencies far (much higher or much lower) from those typically observed could mean a sampling problem. In Figure \ref{fig:histAFP} shows this histogram for the pre-treatment data (first 7 times) of patient ``D'' in the antibiotics study\cite{antibiotic}. On the other hand, we could define the ZRF (Zero Relative Frequency, thereby ranging from 0 to 1) of an element (taxon) as the portion of times where it is zero, i.e., it is not found. Attending to all taxa, we can plot the ZRF histogram, which then lies on the horizontal axis of the plot. The vertical axis shows the number of taxa, so the height of a bar represents the amount of taxa that have determinate ZRF. In this respect, the bar over $0.0$ counts the quantity of taxa that are present at every time point of the data set (aka ``core''), while the bar over $1.0$ would count the total of taxa that are never found (this bar never appears because all these ``null'' elements are automatically filtered by the software. Figure \ref{fig:histZRF} shows this plot for the healthy patient A of the IBS study performed in our lab\cite{IBS}. There, we can see that $12$ taxa are present at all the time points of the time series while $9$ taxa basically appear only once. So, this plot is clearly useful to notice how the ``core'' is distributed.

\begin{figure}		
	\centering		
 	\includegraphics[width=0.8\textwidth]{results/hist/antibiotics_h_patD_pre_AbsFreqPlot}		
 	\caption{Histogram with the absolute frequencies of the pre-treatment data (7 first times) of patient ``D'' in the antibiotics study\cite{antibiotic}. We can see that there are no \emph{outlayers} among the total taxa sum for the time series points, so all of them were considered for further analysis by \CC\ software}		
 	\label{fig:histAFP}		
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/hist/IBS_h_A_amplicons_family_ZRFhist.pdf}
	\caption{Histogram with the relative frequency of zero for the elements (taxa) in the data for patient A of the IBS study\cite{IBS}. In this particular case, the bar over $0.0$ counts the quantity of taxa that are present at every time point of the data set (``core''), while the bar over $1.0$ sums the total of taxa that are never found (this bar never appears because all these ``null'' elements are automatically filtered by \CC): here, $12$ taxa are present at all the time points of the time series while $9$ taxa basically appear only once.}
	\label{fig:histZRF}
\end{figure}

A 2D semi-logarithmic histogram representing deviations from the mean versus the mean itself is a useful tool in the analysis of the stability of ranking processes in complex systems\cite{ranking}. Figure \ref{fig:hist2D} shows this 2D deviation plot for the patient A of the IBS study\cite{IBS}.

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/hist/IBS_h_A_amplicons_family_hist2D.pdf}
	\caption{2D histogram deviation plot of the data for patient A of the IBS study\cite{IBS}}
	\label{fig:hist2D}
\end{figure}

\subsubsection*{Correlation and Rank Plots} 
\CC\ generates two different plots falling under this category, as well as Excel files with the resulting matrices. On the one hand, the elements correlation matrix plot shows a correlation matrix among the taxa, calculated with the time as independent variable. For these calculations, the data set is not normalized to avoid entering an additional constraint. Figure \ref{fig:corrElm} shows this matrix for the most dominant taxa present in the data of the patient ``A'' of the IBS study\cite{IBS}. On the other hand, the rank dynamics and stability plot shows the variation in the rank with time for the most dominant taxa and their calculated RSI, as discussed in Material and Methods. Figure \ref{fig:corrank} shows this plot for the taxa of the aforementioned patient A.

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{results/corrank/IBS_h_A_amplicons_family_ElmCorrelation.pdf}
	\caption{Element correlation plot of the for the most dominant taxa in the data for patient A of the IBS study\cite{IBS}}
	\label{fig:corrElm}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{results/corrank/IBS_h_A_amplicons_family_Rank.pdf}
	\caption{Matrix showing the rank variation throughout time for the most dominant elements (taxa) and their calculated Rank Stability Index (as discussed in Material and Methods) in the data  for patient A of the IBS study\cite{IBS}}
	\label{fig:corrank}
\end{figure}

%%%
\subsection*{Time dependence of model parameters}
